Minicaml, a purely functional, didactical programming language\\WORK IN PROGRESS DRAFT

### 2020-01-16 paper/paper.tex
mutable states in file runner

%% Language and font encodings. This says how to do hyphenation on end of lines.
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{collectbox}
%\usepackage{tgpagella} % text only
%\usepackage{mathpazo}
\usepackage{stix2}

@@

%% Sets page size and margins. You can edit this to your liking
\usepackage[top=1.3cm, bottom=2.0cm, outer=2.5cm, inner=2.5cm, heightrounded,
marginparwidth=1cm, marginparsep=0.4cm, margin=1.2cm]{geometry}

@@

\lstdefinestyle{caml} {language=caml, basicstyle=\ttfamily, columns=[c]fixed,
 keywordstyle=\color{vblue}, identifierstyle=\color{black},
 commentstyle=\color{vgreen}, upquote=true, commentstyle=, breaklines=true,
 showstringspaces=false, stringstyle=\color{blue},
 literate={'"'}{\textquotesingle "\textquotesingle}3}

@@

\begin{abstract}
\textbf{minicaml} is a dinamically typed and purely functional interpreted
programming language. It is based on the Professor Gianluigi Ferrari and
Professor Francesca Levi's minicaml, an evaluation example to show students
attending the Programming 2 course at the University of Pisa how interpreters
work. It is an interpreted language heavily inspired from the OCaml, Haskell and
Scheme languages, with static (lexical scoping), eager and lazy evaluation and a
didactical REPL that shows each AST expression and each evaluation step.
\end{abstract}

@@

\section{Purity Inference}
An important feature of the minicaml language is the purity inference algorithm,
which is performed statically on expressions before evaluation. It is an
interpretation of expressions over the domain of purity, meant to prevent side
effects by signal an error if they are contained inside the programs written in
the language. Expressions are tagged by the algorithm with the \texttt{Pure},
\texttt{Impure} and \texttt{Numerical} labels. An \texttt{Impure} expression is
an expression that contains calls to primitives that perform I/O operations,
mutable variables and/or imperative style assignments. A \texttt{Numerical}
expression is an expression where only numerical operations are performed.
\texttt{Pure} expressions are those which do not fall into the previous two
categories. To perform impure side effects, the programmer has two constructs
available called \textbf{purity blocks}. By default, the evaluator is in an
\texttt{Uncertain} context, which means that it will not allow side effects to
be carried on by evaluation, but will allow evaluating purity blocks that change
the currently allowed purity context. The \texttt{impure} statement takes an
expression (the block) and evaluates it in a context where the allowed purity is
\texttt{Impure}, so that side effects may be performed. The other construct
available, the \texttt{pure} statement, takes an expression and enforces a
\texttt{Pure} context, meaning that side effects and nested impure blocks will
not be allowed inside of the expression.

@@

\section{Evaluation}
\texttt{minicaml}'s evaluator is heavily inspired by the Metacircular Evaluator defined in the highly acclaimed textbook \textit{Structure and Interpretation of Computer Programs} \cite{Abelson1996}
\subsection{Operational Semantics}

@@

\appendix
\clearpage
\section{Parsing Grammar}
\label{grammar}

\lstinputlisting[style=txt]{grammar.txt}

>>>>

%% Language and font encodings. This says how to do hyphenation on end of lines.
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage[title]{appendix}
\usepackage{collectbox}
%\usepackage{tgpagella} % text only
%\usepackage{mathpazo}
\usepackage{stix2}

@@

%% Sets page size and margins. You can edit this to your liking
\usepackage[top=1.3cm, bottom=2.0cm, outer=1.2cm, inner=1.2cm, heightrounded,
marginparwidth=1cm, marginparsep=0.4cm]{geometry}

@@

\lstdefinestyle{caml} {language=caml, basicstyle=\ttfamily, columns=[c]fixed,
 keywordstyle=\color{vblue}, identifierstyle=\color{black},
 commentstyle=\color{vgreen}, upquote=true, commentstyle=, breaklines=true,
 showstringspaces=false, stringstyle=\color{blue}, tabsize=2,
 literate={'"'}{\textquotesingle "\textquotesingle}3}

@@

\begin{abstract}
\textbf{minicaml} is a dyhamically typed and purely functional interpreted
programming language. It is based on the Professor Gianluigi Ferrari and
Professor Francesca Levi's minicaml, an evaluation example to show students
attending the Programming 2 course at the University of Pisa how interpreters
work. It is an interpreted language heavily inspired from the OCaml, Haskell and
Scheme languages, with static (lexical scoping), eager and lazy evaluation and a
didactical REPL that shows each AST expression and each evaluation step.
\end{abstract}

@@

\section{Purity Inference}
An important feature of the minicaml language is the purity inference algorithm,
which is performed statically on expressions before evaluation. It is an interpretation of expressions over the domain of purity, meant to
prevent side effects by signal an error if they are contained inside the
programs written in the language. Expressions are tagged by the algorithm with
the \texttt{Pure}, \texttt{Impure} and \texttt{Numerical} labels. An
\texttt{Impure} expression is an expression that contains calls to primitives
that perform I/O operations, mutable variables and/or imperative style
assignments. A \texttt{Numerical} expression is an expression where only
numerical operations are performed; \texttt{Pure} expressions are those which do not fall into the previous two
categories.

To achieve the execution of impure side effects, the programmer has two constructs
available called \textbf{purity blocks}. By default, the evaluator is in an
\texttt{Uncertain} context, which means that it will not allow side effects to
be carried on by evaluation, but will allow evaluating purity blocks that change
the currently allowed purity context. The \texttt{impure} statement takes an
expression (the block) and evaluates it in a context where the allowed purity is
\texttt{Impure}, so that side effects may be performed. The other construct
available, the \texttt{pure} statement, takes an expression and enforces a
\texttt{Pure} context, meaning that side effects and nested impure blocks will
not be allowed inside of the expression.

@@

\section{Evaluator}
\texttt{minicaml}'s evaluator is heavily inspired by the Metacircular Evaluator defined in the
highly acclaimed textbook \textit{Structure and Interpretation of Computer Programs} \cite{Abelson1996}.

\section{Primitives}
The language primitives that are implemented in OCaml are organized in modules
separated by functionality. Each primitive is a function that accepts a list of
evaluated values and returns a single reduced value; therefore they have a type
of \texttt{evt list -> evt}. OCaml primitives have to perform internal
typechecking and unpacking of the arguments they receive from the minicaml
calls.

From the evaluator's perspective, primitives are organized in a table
such that when a symbol gets evaluated, it is looked up in the primitives table,
if there is a match then the found primitive's name is wrapped in an
\texttt{ApplyPrimitive} expression nestedinside of a lazy lambda expression that
permits partial application. When the evaluator finally encounters an
\texttt{ApplyPrimitive} expression, the primitive OCaml function is extracted,
applied to the arguments and the resulting value is returned by the current
evaluator call. If a primitive is not found when looking up for a symbol, then a
symbol lookup is performed in the current environment.

Some primitives, such as catamorphic procedures, are not native OCaml functions
but small expressions written directly in minicaml; those primitives are kept as
lazy expressions into the same table as native OCaml primitives. The key
difference between the two resides in the fact that those textual minicaml
primitives are not transformed into a function which body contains only an
\texttt{ApplyPrimitive} call, but are instead parsed and analyzed at run time.
The resulting additional startup time caused by parsing and analysis is
proportional to the number of textual form primitives in the table and therefore
quite irrelevant on non-embedded computer systems. The \textit{fold left} and
\textit{fold right} catamorphic primitives are written directly in the minicaml
language and are hereby provided as examples.

\begin{lstlisting}[style=caml,caption=The tail recursive left fold procedure]
fun f z l ->
if typeof l = "list" then
  let aux = fun f z l ->
    if l = [] then z else
	  aux f (f z (head l)) (tail l)
	in aux f z l
else if typeof l = "dict" then
	let aux = fun f z kl vl ->
		if kl = [] && vl = [] then z else
		aux f (f z (head vl)) (tail kl) (tail vl)
	in aux f z (getkeys l) (getvalues l)
else failwith "value is not iterable"
\end{lstlisting}

\begin{lstlisting}[style=caml,caption=The right fold procedure]
fun f z l ->
if typeof l = "list" then
   let aux = fun f z l ->
	  if l = [] then z else
	  f (head l) (aux f z (tail l))
   in aux f z l
else if typeof l = "dict" then
   let aux = fun f z kl vl ->
	  if kl = [] && vl = [] then z else
	  f (head vl) (aux f z (tail kl) (tail vl))
   in aux f z (getkeys l) (getvalues l)
else failwith "value is not iterable"
\end{lstlisting}

@@

\iffalse
\subsection{Operational Semantics}

\fi

\begin{appendices}
  \section{Parsing Grammar}
  \label{grammar}
  \lstinputlisting[style=txt]{grammar.txt}
\end{appendices}
